### 2014-04-02 ApTranscriptome_TR.Rmd
Analysis updated through 'thermal breadth'

```{r expression_shape, eval=TRUE, echo=TRUE}

response.type <- ldply(interaction.transcripts, .progress = "text", .inform = TRUE, RxNtype)

# plot
#qplot(val, pTPM, facets = .~colony, data = A22.pout, geom="smooth")

# merge RxN results with expression values
temperature.transcripts.TPM <- TPM.dt.sub[names(temperature.transcripts)]

# apply the 'RxNply' function to determine maximum and minimum temperatures of expression, as well as the expression level at the optimum temperature of 19.25C
temperature.transcripts.type <- ddply(temperature.transcripts.TPM, .(Transcript), RxNply)
temperature.transcripts.type$max.val <- as.numeric(temperature.transcripts.type$max.val)
temperature.transcripts.type$min.val <- as.numeric(temperature.transcripts.type$min.val)
temperature.transcripts.type$opt.exp <- as.numeric(temperature.transcripts.type$opt.exp)
temperature.transcripts.type$A22.exp.type <- as.factor(temperature.transcripts.type$A22.exp.type)

# repeat for transcripts that have an interaction with colongy
# A22
A22.interaction.transcripts.type <- ddply(interaction.transcripts.TPM["A22"], .(Transcript), RxNply)
A22.interaction.transcripts.type$max.val <- as.numeric(A22.interaction.transcripts.type$max.val)
A22.interaction.transcripts.type$min.val <- as.numeric(A22.interaction.transcripts.type$min.val)
A22.interaction.transcripts.type$opt.exp <- as.numeric(A22.interaction.transcripts.type$opt.exp)
A22.interaction.transcripts.type$A22.exp.type <- as.factor(A22.interaction.transcripts.type$A22.exp.type)
str(A22.interaction.transcripts.type)

# Ar
Ar.interaction.transcripts.type <- ddply(interaction.transcripts.TPM["Ar"], .(Transcript), RxNply)
Ar.interaction.transcripts.type$max.val <- as.numeric(Ar.interaction.transcripts.type$max.val)
Ar.interaction.transcripts.type$min.val <- as.numeric(Ar.interaction.transcripts.type$min.val)
Ar.interaction.transcripts.type$opt.exp <- as.numeric(Ar.interaction.transcripts.type$opt.exp)
Ar.interaction.transcripts.type$A22.exp.type <- as.factor(Ar.interaction.transcripts.type$A22.exp.type)
str(Ar.interaction.transcripts.type)

# combine into single data.table
A22.interaction.transcripts.type$colony <- "A22"
Ar.interaction.transcripts.type$colony <- "Ar"
# first copy of temperature.transcripts.type as colony A22
temperature.transcripts.type$colony <- "A22"
# combine
responsive.transcripts.type <- rbind(A22.interaction.transcripts.type, Ar.interaction.transcripts.type, temperature.transcripts.type)
# add second copy of temperature.transcripts.type as colony Ar
temperature.transcripts.type$colony <- "Ar"
responsive.transcripts.type <- rbind(responsive.transcripts.type, temperature.transcripts.type)
responsive.transcripts.type$colony <- as.factor(responsive.transcripts.type$colony)
str(responsive.transcripts.type)
```

```{r expression_A22.exp.type_table, eval=TRUE, echo=FALSE, results='asis'}
# table type of response
A22.type.table <- table(responsive.transcripts.type[responsive.transcripts.type$colony == "A22", 'A22.exp.type'])
Ar.type.table <- table(responsive.transcripts.type[responsive.transcripts.type$colony == "Ar", 'A22.exp.type'])

A22.exp.type.table <- rbind(A22.type.table, Ar.type.table)
rownames(A22.exp.type.table) <- c("A22", "Ar")

pandoc.table(A22.exp.type.table, style="rmarkdown", caption = "Number of transcripts with maximum expression at high, low, intermediate or both high and low (bimodal) temperatures.")
```

@@

@@

```{r shiny_file}
# scale expression values 
responsive.transcripts.TPM <- TPM.dt.sub[responsive.transcripts]
responsive.transcripts.TPM[,exp.scaled:=scale(TPM), by = Transcript]
str(responsive.transcripts.TPM)
write.csv(responsive.transcripts.TPM, file = "ApRxN-shinyapp/responsive.transcripts.TPM.csv", quote = TRUE, row.names = FALSE)
```

@@

@@

Compare the expression levels at optimum (19.25C) between the two colonies for genes in each expression group. Specifically, are genes that are upregulated at high temperatures in A22 more highly expressed at 21C in Ar? Conversely, are genes that are upregulated at low temps more highly expressed at 21C in A22?

```{r optimum_expression_comparison, cache=TRUE, eval=FALSE}
# list of transcripts that are 'high' expressed in A22
A22_high_transcripts <- responsive.transcripts.type[which(responsive.transcripts.type$colony == "A22" & responsive.transcripts.type$A22.exp.type == "High"), "Transcript"]
# dataframe of transcripts from both colonies that are 'high' expressed in A22
A22_high_df <- responsive.transcripts.type[which(responsive.transcripts.type$Transcript %in% A22_high_transcripts), ]

# Compare expression at optimum temp (19.25C) between colonies using t-test
A22_high_df$colony <- as.factor(A22_high_df$colony)
boxplot(data = A22_high_df, opt.exp ~ colony)
boxplot(data = A22_high_df, log(opt.exp+10) ~ colony)
t.test(A22_high_df[which(A22_high_df$colony == "Ar"), "opt.exp"],  A22_high_df[which(A22_high_df$colony == "A22"), "opt.exp"])

# remove outlier transcripts
A22_high_df_out <- A22_high_df[-which(A22_high_df$opt.exp > 1000), ]
t.test(A22_high_df_out[which(A22_high_df_out$colony == "Ar"), "opt.exp"],  A22_high_df_out[which(A22_high_df_out$colony == "A22"), "opt.exp"])
```

@@

```{r A22_high_wilcoxon}
wilcox.test(A22_high_df[which(A22_high_df$colony == "A22"), "opt.exp"],  A22_high_df[which(A22_high_df$colony == "Ar"), "opt.exp"], alternative = "two.sided", paired = TRUE, conf.int = TRUE)
```

@@

```{r Ar_low_wilcoxon}
# list of transcripts that are 'Low' expressed in Ar
Ar_low_transcripts <- responsive.transcripts.type[which(responsive.transcripts.type$colony == "Ar" & responsive.transcripts.type$A22.exp.type == "Low"), "Transcript"]
# dataframe of transcripts from both colonies that are 'Low' expressed in Ar
Ar_low_df <- responsive.transcripts.type[which(responsive.transcripts.type$Transcript %in% Ar_low_transcripts), ]
dim(Ar_low_df)

# boxplots
boxplot(data = Ar_low_df, opt.exp ~ colony)
boxplot(data = Ar_low_df, log(opt.exp+1) ~ colony)

# Wilcoxon signed rank-test
wilcox.test(Ar_low_df[which(Ar_low_df$colony == "A22"), "opt.exp"],  Ar_low_df[which(Ar_low_df$colony == "Ar"), "opt.exp"], alternative = "two.sided", paired = TRUE, conf.int = TRUE)
```

No difference in transcript expression at 19.25C between colonies for transcripts that are up-regulated at low temperatures in *Ar*.

@@

```{r intermediate_transcripts}
# list of transcripts that are 'Intermediate' expressed in A22
A22_int_transcripts <- responsive.transcripts.type[which(responsive.transcripts.type$colony == "A22" & responsive.transcripts.type$A22.exp.type == "Intermediate"), "Transcript"]
A22_int_df <- responsive.transcripts.type[which(responsive.transcripts.type$Transcript %in% A22_int_transcripts), ]
A22_int_df$colony <- as.factor(A22_int_df$colony)

# Wilcoxon signed rank-test
wilcox.test(A22_int_df[which(A22_int_df$colony == "A22"), "opt.exp"],  A22_int_df[which(A22_int_df$colony == "Ar"), "opt.exp"], alternative = "two.sided", paired = TRUE, conf.int = TRUE)

# list of transcripts that are 'Intermediate' expressed in Ar
Ar_int_transcripts <- responsive.transcripts.type[which(responsive.transcripts.type$colony == "Ar" & responsive.transcripts.type$A22.exp.type == "Intermediate"), "Transcript"]
Ar_int_df <- responsive.transcripts.type[which(responsive.transcripts.type$Transcript %in% Ar_int_transcripts), ]

# Wilcoxon signed rank-test
wilcox.test(Ar_int_df[which(Ar_int_df$colony == "A22"), "opt.exp"],  Ar_int_df[which(Ar_int_df$colony == "Ar"), "opt.exp"], alternative = "two.sided", paired = TRUE, conf.int = TRUE)
```

As expected, for "Intermediate" expressed transcripts there is no difference in expression at 19.25C between colonies.

@@

```{r calc_Intermediate_variance}
# extract 'Intermediate' expressed transcripts for A22 colony
setkey(TPM.dt.sub, Transcript)
A22trlist <- A22.interaction.transcripts.type[which(A22.interaction.transcripts.type$A22.exp.type == "Intermediate"), "Transcript"]
A22.TPM.int.dt <- TPM.dt.sub[A22trlist]
setkey(A22.TPM.int.dt, colony)
A22.TPM.int.dt <- A22.TPM.int.dt["A22"]
str(A22.TPM.int.dt)

# calculate standard deviation of expression for each transcript
A22_int_sd <- ddply(A22.TPM.int.dt, .(Transcript), RxNsd.concave)

# repeat for Ar
Artrlist <- Ar.interaction.transcripts.type[which(Ar.interaction.transcripts.type$A22.exp.type == "Intermediate"), "Transcript"]
Ar.TPM.int.dt <- TPM.dt.sub[Artrlist]
setkey(Ar.TPM.int.dt, colony)
Ar.TPM.int.dt <- Ar.TPM.int.dt["Ar"]
str(Ar.TPM.int.dt)

Ar_int_sd <- ddply(Ar.TPM.int.dt, .(Transcript), RxNsd.concave)
```

With these values calculated, I compare the standard deviation of expression between colonies.

```{r compare_Intermediate_variance}
# T-test
t.test(A22_int_sd$exp_sd, Ar_int_sd$exp_sd)

# Plot
# prep data
A22_int_var$colony <- "A22"
Ar_int_var$colony <- "Ar"
comb_int_var <- rbind(A22_int_var, Ar_int_var)
# ggplot
g1 <- ggplot(comb_int_var, aes(x=exp_sd, fill=colony)) + geom_density(alpha=0.2, position="identity")
g1 + scale_y_continuous(name="Density") +
  scale_x_continuous(name=expression("Standard deviation of expression"))
```

@@

>>>>

```{r expression_shape, eval=TRUE, echo=TRUE}
# calculate response type for transcripts with interactions
interaction.response.type <- ldply(interaction.transcripts, .progress = "text", .inform = TRUE, RxNtype)
stopifnot(nrow(interaction.response.type) == length(interaction.transcripts))

# calculate response types for transcripts without interactions
temperature.response.type <- ldply(temperature.transcripts, .progress = "text", .inform = TRUE, RxNtype)
stopifnot(nrow(temperature.response.type) == length(temperature.transcripts))

# merge results
Ap.response.type <- rbind(interaction.response.type, temperature.response.type)
colnames(Ap.response.type)[which(colnames(Ap.response.type) == ".id")] <- "Transcript"
str(Ap.response.type)
```

Examine transcripts by temperature response. Compare between colonies.

```{r temperature_response, eval=TRUE, echo=FALSE, results='asis'}
# table type of response
A22.type.table <- table(Ap.response.type[ , 'A22.type'])
Ar.type.table <- table(Ap.response.type[ , 'Ar.type'])

Ap.type.table <- rbind(A22.type.table, Ar.type.table)
rownames(Ap.type.table) <- c("A22", "Ar")

pandoc.table(Ap.type.table, style="rmarkdown", caption = "Number of transcripts with maximum expression at high, low, intermediate or both high and low (bimodal) temperatures.")
```

@@

**2014-04-02 -- CURRENTLY BROKEN**

@@

```{r shiny_file, eval=FALSE, echo=FALSE}
shiny.transcripts <- TPM.dt.sub[Ap.response.type$Transcript]

# scale expression values 
shiny.transcripts[,pTPM.scaled:=scale(pTPM), by = Transcript]
str(responsive.transcripts.TPM)
write.csv(responsive.transcripts.TPM, file = "ApRxN-shinyapp/responsive.transcripts.TPM.csv", quote = TRUE, row.names = FALSE)
```

@@

In this section, I compare the thermal reactionome between the *Ar* and *A22* colonies.

@@

Selection may have acted in response to thermal stress such that some genes are constitutively activated in one colony while plastically-expressed in the other. I evaluate this hypothesis by comparing expression levels at the at optimum (19.5C) temperature between the two colonies for genes in that are either in the 'High' or 'Low' expression group in the other colony. Specifically, I expect that genes upregulated at high temperatures in *A22* are more highly expressed at 19.5C in *Ar*? Conversely, I expect genes that are upregulated at low temps in *Ar* to have greater constitutive expression in *A22*?

```{r optimum_expression_comparison, cache=TRUE, eval=FALSE}
# list of transcripts that are 'high' expressed in A22
A22.high.transcripts <- Ap.response.type[which(Ap.response.type$A22.type == "High"), ]

# Compare expression at optimum temp (A22.opt) between colonies using t-test
boxplot(A22.high.transcripts$A22.opt, A22.high.transcripts$Ar.opt)
boxplot(log(A22.high.transcripts$A22.opt+1), log(A22.high.transcripts$Ar.opt+1))

# T test
t.test(log(A22.high.transcripts$A22.opt+1), log(A22.high.transcripts$Ar.opt+1))
```

@@

```{r A22_high_wilcoxon}
wilcox.test(A22.high.transcripts$A22.opt, A22.high.transcripts$Ar.opt, alternative = "two.sided", paired = TRUE, conf.int = TRUE)
```

@@

```{r Ar_low_wilcoxon}
# list of transcripts that are 'high' expressed in Ar
Ar.low.transcripts <- Ap.response.type[which(Ap.response.type$Ar.type == "Low"), ]

# Compare expression at optimum temp (Ar.opt) between colonies using t-test
boxplot(Ar.low.transcripts$A22.opt, Ar.low.transcripts$Ar.opt)
boxplot(log(Ar.low.transcripts$A22.opt+1), log(Ar.low.transcripts$Ar.opt+1))

# T test
t.test(log(Ar.low.transcripts$A22.opt+1), log(Ar.low.transcripts$Ar.opt+1))

# Wilcoxon signed rank-test
wilcox.test(Ar.low.transcripts$A22.opt, Ar.low.transcripts$Ar.opt, alternative = "two.sided", paired = TRUE, conf.int = TRUE)
```

Counter to expectations, expression at 19.5C is also greater in *Ar* than *A22* for transcripts upregulated at low temperatures in *Ar*.

@@

```{r intermediate_transcripts}
# list of transcripts that are 'Intermediate' expressed in Ar
Ar.int.transcripts <- Ap.response.type[which(Ap.response.type$Ar.type == "Intermediate"), ]
# T test
t.test(log(Ar.int.transcripts$A22.opt+1), log(Ar.int.transcripts$Ar.opt+1))
# Wilcoxon signed rank-test
wilcox.test(Ar.int.transcripts$A22.opt, Ar.int.transcripts$Ar.opt, alternative = "two.sided", paired = TRUE, conf.int = TRUE)

# list of transcripts that are 'Intermediate' expressed in A22
A22.int.transcripts <- Ap.response.type[which(Ap.response.type$A22.type == "Intermediate"), ]
# T test
t.test(log(A22.int.transcripts$A22.opt+1), log(A22.int.transcripts$Ar.opt+1))
# Wilcoxon signed rank-test
wilcox.test(A22.int.transcripts$A22.opt, A22.int.transcripts$Ar.opt, alternative = "two.sided", paired = TRUE, conf.int = TRUE)
```

One of two tests, for *Ar* intermediate transcripts also finds greater constitutive expression in *Ar* than *A22*.

@@

```{r thermal_breadth}
# extract 'Intermediate' expressed transcripts for A22 colony
A22.int.lm <- RxNlmAIC[A22.int.transcripts$Transcript]
length(A22.int.lm)

# calculate standard deviation of expression for each transcript
transcriptVar <- function(lmitem, colony) {
    # predict only for A22
    pred.vals <- seq(from = 0, to = 38.5, by = 0.5)
    newdf <- data.frame(val = pred.vals, colony = rep(colony, length = length(pred.vals)))
    pout <- predict(lmitem, newdata=newdf)
    pout <- cbind(newdf, pTPM = pout)

    # prediction can give values very close to zero or negative. biologically, these are meaningless so change to zero
    pout$pTPM <- ifelse(pout$pTPM < 0, 0, round(pout$pTPM, 5))

    # draw 1000 'temps' weighted by predicted expression
    random.draw <- sample(size = 1000, pred.vals, prob = pout$pTPM, replace = TRUE)

    # calculate standard deviation of weighted random.draw
    data.frame(exp_sd = sd(random.draw))
    } # end function

# apply function to all transcripts
A22.int.sd <- unlist(Map(transcriptVar, A22.int.lm, colony = "A22"))

# repeat for Ar
Ar.int.lm <- RxNlmAIC[Ar.int.transcripts$Transcript]
Ar.int.sd <- unlist(Map(transcriptVar, Ar.int.lm, colony = "Ar"))

# T-test comparing standard deviation of expression between colonies
t.test(A22.int.sd, Ar.int.sd)

```{r plotthermalbreadth, echo=FALSE, eval=TRUE)
# prep data for plotting
A22.int.sd <- data.frame(colony = "A22", exp_sd = A22.int.sd)
Ar.int.sd <- data.frame(colony = "Ar", exp_sd = Ar.int.sd)
Ap.int.sd <- rbind(A22.int.sd, Ar.int.sd)

# ggplot
g1 <- ggplot(Ap.int.sd, aes(x=exp_sd, fill=colony)) + geom_density(alpha=0.2, position="identity")
g1 + scale_y_continuous(name="Density") +
  scale_x_continuous(name=expression("Standard deviation of expression"))
```

@@

~~Continue from here~~
